Weighted Model Counting Without Parameter Variables

### 2021-03-03 doc/paper3/paper.tex
Formulate the correctness theroem for MCWMC

\crefname{enumii}{Condition}{Conditions}

@@

TODO: Mention that formulas, clauses, and models are all treated as sets.

@@

\begin{algorithm}[t]
  \caption{WMC to PBP transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{indicator variable $v \in X_I$ \label{line:foreach2start}}{
    \If{$\{[v]_1^p, [\neg v]_1^q\} \subseteq F$ for some $p$ and $q$}{
      $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
      \}$\; \label{line:foreach2end}
    }
  }
  replace every variable $v$ in $F$ with $\rename{$v$}$\;
  \Return{$(F, X_I, \omega)$}\;
  \Fn{\rename{$v$}}{
    $S \gets \{u \in X_P \mid u \le v\}$\;
    \lIf{$S = \emptyset$}{\Return{$v$}}
    \Return{$v - d(\max S)$}\;
  }
\end{algorithm}

@@

\paragraph{TODO}
\begin{itemize}
\item check if each condition is actually used. Maybe turn this into a paragraph
  that gives an overview of the proof.
\item condition 1 is necessary in both cases because we're ignoring the weights
  of indicator variables (explicitly acknowledge this)
\item add the \texttt{cd05}/\texttt{cd06} correctness theorem
\end{itemize}

@@

\begin{theorem}[Correctness] \label{thm:correctness}
  \Cref{alg:transformation}, when given a WMC instance $(\phi, X_I, X_P,
  w)$, returns PBP instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$,
      \item there is no clause $c \in \phi$ such that $\neg p \in c$,
      \item if $\{ p \} \in \phi$, then there is no clause $c \in \phi$ such
        that $c \ne \{ p \}$ and $p \in c$, \label{cond:just_parameter}
        % 'prime implicants are logically disjoint'
      \item and for any $c, d \in \phi$ such that $c \ne d$, $p \in c$ and $p
        \in d$, $\bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l
          \in d \setminus \{ p \}} \neg l$ is false. \label{cond:disjoint}
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem}
\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc2}
  \end{equation}
  is a PBP instance with the same answer as the given WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \left( \prod_{c \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \eqref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

@@

\section{Experimental Evaluation (mostly done)}
% TODO: what are the encodings compiled with?

>>>>

\crefname{enumi}{Condition}{Conditions}
\crefname{enumii}{Condition}{Conditions}

@@

\paragraph{TODO}
\begin{itemize}
\item Explain two meanings of entailment in the definition of WMC.
\end{itemize}

Throughout the paper, we use set theoretic notation for many concepts in logic.
We identify a model with a set of variables that correspond to the positive
literals in the model (and all other variables are the negative literals of the
model). A formula in CNF is a set of clauses that are part of an implicit
conjunction. Similarly, a clause is a set of literals that are part of an
implicit disjunction. For example, let $\phi = (\neg a \lor b) \land a$ be a
propositional formula over variables $a$ and $b$. Then an equivalent
set-theoretic representation of $\phi$ is $\{ \{ \neg a, b \}, \{ a \} \}$. Any
subset of $\{ a, b \}$ is an interpretation of $\phi$, e.g., $\{ a, b \}$ is a
model of $\phi$ (written $\{ a, b \} \models \phi$), while $\emptyset$ is an
interpretation but not a model.

@@

\begin{algorithm}[t]
  \caption{WMC to PBP transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC (or minimum-cardinality WMC) instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{indicator variable $v \in X_I$ \label{line:foreach2start}}{
    \If{$\{[v]_1^p, [\neg v]_1^q\} \subseteq F$ for some $p$ and $q$}{
      $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
      \}$\; \label{line:foreach2end}
    }
  }
  replace every variable $v$ in $F$ with $\rename{$v$}$\;
  \Return{$(F, X_I, \omega)$}\;
  \Fn{\rename{$v$}}{
    $S \gets \{u \in X_P \mid u \le v\}$\;
    \lIf{$S = \emptyset$}{\Return{$v$}}
    \Return{$v - d(\max S)$}\;
  }
\end{algorithm}

@@

\paragraph{TODO}
\begin{itemize}
\item check if each condition is actually used. Maybe turn this into a paragraph
  that gives an overview of the proof.
\item condition 1 is necessary in both cases because we're ignoring the weights
  of indicator variables (explicitly acknowledge this)
\end{itemize}

@@

\begin{theorem}[Correctness for WMC] \label{thm:correctness}
  \Cref{alg:transformation}, when given a WMC instance $(\phi, X_I, X_P,
  w)$, returns PBP instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) =
    1$, \label{cond:one}
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$, \label{cond:2b2}
      \item there is no clause $c \in \phi$ such that $\neg p \in
        c$, \label{cond:2b3}
      \item if $\{ p \} \in \phi$, then there is no clause $c \in \phi$ such
        that $c \ne \{ p \}$ and $p \in c$, \label{cond:just_parameter}
        % 'prime implicants are logically disjoint'
      \item and for any $c, d \in \phi$ such that $c \ne d$, $p \in c$ and $p
        \in d$, $\bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l
          \in d \setminus \{ p \}} \neg l$ is false. \label{cond:disjoint}
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem}
\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc2}
  \end{equation}
  is a PBP instance with the same answer as the given WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \left( \prod_{c \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \eqref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

@@

\begin{theorem}[Correctness for Minimum-Cardinality WMC]
  Let $(\phi, X_I, X_P, w)$ be a minimum-cardinality WMC instance that satisfies
  \cref{cond:one,cond:2b2,cond:2b3,cond:just_parameter,cond:disjoint} of
  \cref{thm:correctness} as well as the following:
  \begin{enumerate}
  \item for all parameter variables $p \in X_P$, $w(\neg p) = 1$.
  \item all models of $\{ c \in \phi \mid c \cap X_P = \emptyset \}$ (as subsets
    of $X_I$) have the same cardinality;
  \item $\min_{Z \subseteq X_P} |Z|$ such that $Y \cup Z \models \phi$ is the
    same for all $Y \models \{ c \in \phi \mid c \cap X_P = \emptyset \}$.
  \end{enumerate}
  Then \cref{alg:transformation}, when applied to $(\phi, X_I, X_P, w)$, outputs
  a PBP instance with the same answer.
\end{theorem}
\begin{proof} % TODO: introduce the variables that come from the algorithm
  We split the proof into two parts. Let
  \[
    F = \left\{ [c]_0^1 \mid c \in \phi \text{, } c \cap X_P = \emptyset
    \right\} \cup \left\{ \left[ \bigwedge_{l \in c \setminus \{ p \}} \neg l
      \right]_1^{w(p)} \;\middle|\; p \in X_P \text{, } p \in c \in \phi
      \text{, } c \ne \{ p \} \right\}
  \]
  be as returned by \cref{alg:transformation}. In the first part, we show that
  there is a bijection between minimum-cardinality models of $\phi$ and $Y
  \subseteq X_I$ such that $\left(\prod_{f \in F} f\right)(Y) \ne 0$ (with the
  assumption that $w(p) \ne 0$ for all $p \in X_P$).\footnote{The assumption is
    only there to make the proof more convenient and can be omitted by
    reformulating the proof in a slightly more cumbersome way.} Let $Y \subseteq
  X_I$ and $Y' \subseteq X_I \cup X_P$ be related via this bijection. Then in
  the second part we will show that
  \[
    \prod_{Y' \models l} w(l) = \left(\prod_{f \in F} f\right)(Y).
  \]

  On the one hand, if $Y \subseteq X_I \cup X_P$ is a minimum-cardinality model
  of $\phi$, then
  \[
    \left(\prod_{f \in F}\right)(Y \cap X_I) \ne 0
  \]
  under the given assumptions. On the other hand, if $Y \subseteq X_I$ is such
  that
  \[
    \left(\prod_{f \in F}\right)(Y) \ne 0,
  \]
  then $Y \models \{ c \in \phi \mid c \cap X_P = \emptyset \}$. Let $Y
  \subseteq Z \subseteq X_I \cup X_P$ be the smallest superset of $Y$ such that
  $Z \models \phi$ (it exists by \cref{cond:2b3} of \cref{thm:correctness}). We
  need to show that $Z$ is minimum cardinality.
\end{proof} % TODO: first part
% TODO: make a separate definition for minimum cardinality model and use it in
% the definition of MCWMC
% TODO: use the MCWMC abbreviation
% TODO: my current proof doesn't use omega. Maybe just say that the proof is
% easily extendable to phi s.t. omega != 1 similarly to the proof of
% thm:correctness.

\section{Experimental Evaluation (mostly done)}
% TODO: define cardinality of model
% TODO: what are the encodings compiled with?
